---
title: "AI-Assisted Statistical Analysis in Graduate Epidemiology Education"
subtitle: "Integrating Claude with RStudio for Teaching Regression Methods"
author: "Muntasir Masum, PhD"
institute: "Department of Epidemiology & Biostatistics<br>University at Albany, SUNY<br><br>[AI+ Annual Symposium 2026](https://events.albany.edu/event/ai-plus-symposium)"
date: "`r Sys.Date()`"
format:
  revealjs:
    theme: ["simple", ai-epi-workshop.scss]
    slide-number: c/t
    width: 1600
    height: 900
    footer: "AI Plus Annual Symposium 2026"
    chalkboard:
      buttons: false
    preview-links: auto
    echo: true
    mouse-wheel: true
---

```{r}
#| label: setup
#| include: false

# Load required packages
library(tidyverse)
library(NHANES)
library(broom)
library(knitr)
library(kableExtra)
library(plotly)

# Set global chunk options
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 10,
  fig.height = 5,
  cache = FALSE
)

# Set ggplot theme
theme_set(theme_minimal(base_size = 16))
```

## Welcome

### Today's Focus

::: incremental
- See Claude + RStudio in action
- Real statistical workflows
- Practical demonstrations
- Q&A and discussion
:::

## Workshop Overview

### 75-Minute Structure

::: columns
::: {.column width="33%"}
**Part 1: Context** (15 min)

- Teaching challenge
- Why AI + R?
- Course integration
:::

::: {.column width="33%"}
**Part 2: Demonstrations** (45 min)

- Coding workflow
- Analysis workflow
- Visualization
- Debugging
:::

::: {.column width="33%"}
**Part 3: Discussion** (15 min)

- Implementation
- Academic integrity
- Q&A
:::
:::

::: footer
Materials available at: [GitHub repository link]
:::

## Who Is This For?

::: incremental
- **Educators** teaching quantitative methods
- **Students** learning R and statistics
- **Researchers** working with data
- **Anyone** interested in AI-assisted analysis
:::

. . .

**No assumptions about:** R experience • Statistics background • Epidemiology knowledge

## The Teaching Challenge

**Students come from diverse backgrounds**

- Biology, Psychology, Public Health, Policy
- Variable quantitative preparation
- Different comfort with programming

. . .

**Traditional barriers**

- Syntax frustration overshadows statistical thinking
- Limited support outside office hours
- Steep learning curve for R
- Time debugging instead of learning

. . .

**The real goal:** Develop analytical thinkers, not code executors

## Why Integrate AI Now?

::: columns
::: {.column width="50%"}
**Traditional Approach**

- Lectures + textbook
- Office hours
- TA support (limited)
- Peer learning groups
- Steep learning curve
:::

::: {.column width="50%"}
**AI-Enhanced Approach**

- On-demand explanation
- Contextual guidance
- Code debugging help
- Iterative learning
- Focus on interpretation
:::
:::

## Course Context: EPI 553

**Principles of Statistical Inference II** (Graduate level)

::: columns
::: {.column width="50%"}
**Course Details**

- 12-week semester
- Multiple regression methods
- All analyses in R
- Team research projects
- Poster presentation
:::

::: {.column width="50%"}
**Progressive Integration**

- Weeks 1-4: Structured prompts
- Weeks 5-8: Semi-structured
- Weeks 9-12: Open-ended projects

*Scaffolding → Independence*
:::
:::

## Four Key Principles

1. **Scaffold, Don't Replace** — AI guides, students demonstrate understanding

2. **Context Matters** — Statistical output → Public health implications

3. **Encourage Curiosity** — AI asks questions that deepen learning

4. **Build Independence** — Progressive responsibility transfer

. . .

**Core Philosophy:** AI is a teaching assistant, not a replacement for learning

## Three Ways to Integrate Claude + RStudio

### Choose Your Workflow

::: columns
::: {.column width="33%"}
**1. ClaudeR Package**

- Direct RStudio integration
- Chat interface in RStudio
- Access to R environment
- *Most accessible*

:::

::: {.column width="33%"}
**2. Claude Chat**

- Web browser + RStudio
- Copy/paste workflow
- Manual code execution
- *Intermediate approach*

:::

::: {.column width="33%"}
**3. Claude Code**

- Full AI IDE integration
- Autonomous execution
- Advanced workflows
- *Most powerful*

:::
:::

. . .

**Today's focus:** All three approaches, with emphasis on **ClaudeR** for accessibility

## ClaudeR Package Demo

### What is ClaudeR?

**Direct integration between RStudio and Claude AI**

::: incremental
- R package that connects to Claude Desktop/Code
- Chat interface appears in RStudio
- Claude can "see" your R environment
- Interactive data analysis without leaving RStudio
- **Perfect for students new to AI workflows**
:::

. . .

**Installation:** `devtools::install_github("IMNMV/ClaudeR")`

**GitHub:** [github.com/IMNMV/ClaudeR](https://github.com/IMNMV/ClaudeR)

## ClaudeR in Action

### Live Demonstration

```r
# Start ClaudeR connection
library(ClaudeR)
claudeAddin()

# Now you can chat with Claude directly in RStudio!
# Claude can see your loaded datasets
# Ask questions about your data
# Get help with analysis
```

. . .

**Example conversation:**

::: {style="background-color: #f0f0f0; padding: 20px; border-radius: 10px; margin-top: 20px;"}
👤 **Student:** "I have the NHANES data loaded. Can you help me run a basic summary of systolic blood pressure?"

🤖 **Claude:** "I can see you have NHANES loaded in your environment. Let me help you..."
:::

## Why ClaudeR Matters for Teaching

::: incremental
- **Lower barrier to entry** — Students don't need to learn new tools
- **Familiar environment** — Works within RStudio they already use
- **Context awareness** — Claude sees loaded data and objects
- **Progressive learning** — Students can transition to more advanced workflows
- **Reduces friction** — No switching between browser and IDE
:::

. . .

**This is the approach I recommend for getting started!**

## Today's Research Question

> *Does alcohol consumption predict blood pressure?*

**Dataset:** NHANES 2017-2018

- Outcome: Systolic blood pressure (mmHg)
- Exposure: Alcohol drinks per week
- Covariates: Age, BMI, smoking status, sex

. . .

**Four workflows:** Coding • Analysis • Plotting • Debugging

## Student Question: Coding

*"I need to prepare my NHANES data for regression analysis. I want adults aged 20-65, exclude missing values, and create a drinks-per-week variable. How do I do this in R?"*

. . .

**How Claude helps:**

- Suggests appropriate functions
- Explains data manipulation logic
- Provides annotated code
- Anticipates common pitfalls

## Claude's Approach

**Understanding the task:**

1. Filter age range (20-65)
2. Select relevant variables
3. Handle missing data
4. Create derived variables

## The Code

```{r}
#| eval: false
#| echo: true

# Load packages
library(tidyverse)

# Load NHANES data
data("NHANES")

# Prepare analysis dataset
analysis_data <- NHANES %>%
  filter(Age >= 20, Age <= 65) %>%
  select(ID, Gender, Age, BMI, AlcoholDay, SmokeNow, BPSysAve) %>%
  na.omit() %>%
  mutate(
    drinks_per_week = AlcoholDay * 7,
    current_smoker = ifelse(SmokeNow == "Yes", 1, 0)
  )
```

## Live Demonstration

```{r}
#| label: data-prep
#| echo: true

# Load NHANES data
data("NHANES")

# Prepare analysis dataset
analysis_data <- NHANES %>%
  filter(Age >= 20, Age <= 65) %>%
  filter(!is.na(BPSysAve), !is.na(AlcoholDay), !is.na(BMI),
         !is.na(SmokeNow), !is.na(Gender)) %>%
  mutate(
    drinks_per_week = AlcoholDay * 7,
    current_smoker = ifelse(SmokeNow == "Yes", 1, 0)
  ) %>%
  select(ID, Gender, Age, BMI, drinks_per_week, current_smoker, BPSysAve) %>%
  distinct(ID, .keep_all = TRUE)
```

## Check the Result

```{r}
#| echo: true
glimpse(analysis_data)
```

## Key Takeaways

1. Claude explains **logic** before showing code
2. Code is **annotated** for learning
3. Line-by-line highlighting helps follow flow
4. Student **understands** pipeline, not just copies

. . .

**Teaching moment:** Why filter missing values *before* creating new variables?

## Student Question: Analysis

*"I need to run a multiple regression with blood pressure as outcome and alcohol, age, BMI, smoking, and sex as predictors. How do I interpret the results?"*

## Fitting the Model

```{r}
#| label: model1
#| echo: true

# Fit multiple linear regression
model1 <- lm(BPSysAve ~ drinks_per_week + Age + BMI +
                        current_smoker + Gender,
             data = analysis_data)

# View results
summary(model1)
```

## Claude's Interpretation

**1. Coefficient for alcohol:**

```{r}
#| echo: true
coef(model1)["drinks_per_week"]
```

Each additional drink per week is associated with a `r round(coef(model1)["drinks_per_week"], 3)` mmHg change in systolic BP, holding other factors constant.

## Statistical Significance

**2. P-value:**

```{r}
#| echo: true
summary(model1)$coefficients["drinks_per_week", "Pr(>|t|)"]
```

. . .

**Claude asks:** "What does p = `r format.pval(summary(model1)$coefficients['drinks_per_week', 'Pr(>|t|)'], digits=3)` tell us about this relationship?"

. . .

**Claude asks:** "Is this effect size clinically meaningful for public health?"

## Student Question: Visualization

*"I want to visualize the relationship between alcohol consumption and blood pressure, accounting for sex differences. What's the best plot?"*

. . .

**Claude suggests:**

- Scatter plot with regression lines
- Separate by sex
- Add confidence intervals
- Clear labels

## Building the Plot

```{r}
#| label: vis-demo-code
#| echo: true
#| eval: false

# Create interactive plot with plotly
p <- ggplot(analysis_data, aes(x = drinks_per_week, y = BPSysAve, color = Gender)) +
  geom_point(alpha = 0.3, size = 2) +
  geom_smooth(method = "lm", se = TRUE, linewidth = 1.5) +
  labs(
    title = "Alcohol Consumption and Systolic Blood Pressure",
    subtitle = "NHANES 2017-2018, Adults Aged 20-65",
    x = "Drinks Per Week",
    y = "Systolic Blood Pressure (mmHg)",
    color = "Sex"
  ) +
  theme_minimal(base_size = 18) +
  theme(legend.position = "bottom")

# Convert to interactive plot
ggplotly(p)
```

## Interactive Visualization

```{r}
#| label: vis-demo
#| echo: false
#| fig-width: 10
#| fig-height: 5.5

# Create the plot
p <- ggplot(analysis_data, aes(x = drinks_per_week, y = BPSysAve, color = Gender)) +
  geom_point(alpha = 0.3, size = 2) +
  geom_smooth(method = "lm", se = TRUE, linewidth = 1.5) +
  labs(
    title = "Alcohol Consumption and Systolic Blood Pressure",
    subtitle = "NHANES 2017-2018, Adults Aged 20-65",
    x = "Drinks Per Week",
    y = "Systolic Blood Pressure (mmHg)",
    color = "Sex"
  ) +
  theme_minimal(base_size = 18) +
  theme(legend.position = "bottom")

# Convert to interactive plotly
ggplotly(p, tooltip = c("x", "y", "color")) %>%
  layout(hovermode = "closest")
```

::: {.notes}
Hover over points to see values. Click legend to show/hide groups. Interactive plots help students explore data patterns.
:::

## Iterative Refinement

**Student:** "Can we make this publication-ready?"

. . .

**Claude suggests:**

- Larger text for readability
- Color-blind friendly palette
- Add sample size annotations
- Export at high resolution

## Student Question: Debugging

**Common errors students encounter:**

1. **Missing package errors**
   ```r
   Error: could not find function "ggplot"
   ```

2. **Variable name typos**
   ```r
   Error: object 'drinks_perweek' not found
   ```

3. **Formula syntax errors**
   ```r
   Error in lm.fit(x, y, ...)
   ```

## Claude's Diagnostic Process

**Student shares error:**
```r
Error: object 'current_smoker' not found
```

. . .

**Claude's approach:**

1. Identify the issue (variable doesn't exist)
2. Explain why (not created or misspelled)
3. Suggest fix (check mutate step)
4. Teach pattern (how to verify variables)

## Live Debugging

**Intentional error:**

```{r}
#| error: true
#| echo: true
model_error <- lm(BP_Systolic ~ alcohol_weekly + Age,
                  data = analysis_data)
```

## Claude's Fix

"I see two issues:

1. `BP_Systolic` should be `BPSysAve`
2. `alcohol_weekly` should be `drinks_per_week`"

. . .

**Corrected:**

```{r}
#| echo: true
model_correct <- lm(BPSysAve ~ drinks_per_week + Age,
                    data = analysis_data)
```

## Optional Follow-Along

**For participants:**

1. Install R and RStudio (latest versions)
2. Create Claude account (free at claude.ai)
3. Install packages:
   ```r
   install.packages(c("tidyverse", "NHANES", "broom"))
   ```

**Workshop materials:** GitHub repository [link to be added]

## Academic Integrity

::: columns
::: {.column width="50%"}
**Clear Guidelines**

- AI assistance allowed
- Must cite AI use
- Understand all code
- No copy-paste only
:::

::: {.column width="50%"}
**Assessment Design**

- Timed exams (no AI)
- Interpretation focus
- Code annotation
- Learning reflection
:::
:::

. . .

**Key message:** AI is a learning tool, not a replacement for understanding

## Adapting to Your Context

- **Different disciplines** — Principles apply beyond epidemiology
- **Various skill levels** — Scale scaffolding to student needs
- **Different software** — Concepts transfer to Python, Stata, SAS
- **Online/in-person** — Works in both modalities
- **Class size** — Scalable from seminars to large lectures

## Benefits Observed

::: columns
::: {.column width="50%"}
**For Students**

- Reduced frustration
- Deeper understanding
- More interpretation time
- Confidence building
- 24/7 support access
:::

::: {.column width="50%"}
**For Instructors**

- Focus on concepts
- Better class time use
- Fewer basic questions
- Advanced discussions
- Scalable support
:::
:::

## Challenges & Solutions

| Challenge | Solution |
|-----------|----------|
| Over-reliance on AI | Progressive scaffolding; AI-free assessments |
| Incorrect suggestions | Teach verification; show error examples |
| Equity concerns | Free tier; computer lab access |
| Faculty learning curve | Start small; build gradually |
| Academic integrity | Clear policies; reflection assignments |

## Discussion

**Let's discuss:**

- What challenges do you face in teaching quantitative methods?
- How might AI integration work in your context?
- What concerns do you have?
- What additional support would help?

## Resources

**Workshop Materials**

- GitHub: [repository link]
- Slides: [link to HTML]
- Contact: mmasum@albany.edu

**Additional Resources**

- Quarto: <https://quarto.org>
- Claude: <https://claude.ai>
- NHANES: <https://www.cdc.gov/nchs/nhanes/>

## Thank You!

<br><br>

### Questions?

<br>

**Muntasir Masum, PhD**

mmasum@albany.edu

Department of Epidemiology & Biostatistics

University at Albany, SUNY
